Learning a functional distance between organizations with unsupervised embeddings of scientific mobility

### 2020-07-06 paper/current/main.tex
minor typo
minor change

Human mobility and migration drives major social phenomenon such as the growth and evolution of cities~\autocite{bettencourt2007cities, bettencourt2013cities, wef2017migration, curiel2018citygravity},  epidemics~\autocite{kraemer2020covid, truscott2012epidemicgravity, xia2005measlesgravity, hufnagel2004epidemics, eubank2004outbreaks}, economies~\autocite{kaluza2010cargo, kerr2011immigration, taylor2016refugees, pappalardo2015mobility}, innovation~\autocite{braunerhjelm2020labor, kaiser2018innovation, armano2017innovation, liu2010spillovers},  and scientific research~\autocite{sugimoto2017mostimpact, petersen2018multiscale, jonkers2013return, franzoni2014advantage, morgan2018prestige, rodrigues2016mobility}. 
Common models of mobility, such as the Gravity Model~\autocite{zipf1946gravity, xia2005measlesgravity, jung2008highwaygravity, barbosa2018human, curiel2018citygravity, lewer2008immigrationgravity, hong2016busgravity, truscott2012epidemicgravity, xia2005measlesgravity, simini2012universal}, describes mobility using the \textit{distance} between places.
However, amid a globalized world~\autocite{czaika2018migration, czaika2018globalisation, buchan2009globalization}, geographic distance has less relevance for understanding mobility, whereas other abstract notions of distance~\autocite{boschma2005proximity} pose practical and theoretical challenges.
Using the case of scientific mobility, we leverage the neural embedding model~\autocite{mikolov2013word2vec} to demonstrate a data-driven approach for learning a vector-space embedding that encodes a functional distance~\autocite{brown1970migration, brown1970functional, kim2018functional} between organizations and captures the latent and multi-faceted structure of human mobility.
We train neural embeddings of scientific organizations using mobility trajectories derived from the papers published between 2008 and 2019 of disambiguated authors from the Web of Science database.
Functional distnace is derived as the cosine distance between vectors in the resulting embedding, and is shown to better explain and predict real mobility between organizations than geographic distance. 
Exploiting their unique properties, we interpret the embedding space and observe how it captures latent factors related to culture, language, and prestige of organizations at multiple levels of granularity.
This approach can  learn scalable, dense, and meaningful vector representations of scientific mobility directly from the data, and can easily be generalized to other mobility data.
\end{abstract}

@@

How far apart are two places?
The question is surprisingly hard to answer when it involves human mobility.
While geographic distance may have driven human movements in past eras, it holds less importance to a world connected by rapid transit, complex international relations, and global trade.
For example, a person currently living in Australia is more likely to migrate to the United Kingdom, a country with similar language and culture, than to closer and more populous countries like Indonesia. 
Similarly, a prospective student living in South Korea is more likely to attend university in Canada rather than the closer though diplomatically isolated North Korea. 
Despite its complexity, ubiquitious models of human mobility such as the Gravity~\autocite{zipf1946gravity} and Radiation~\autocite{simini2012universal} model continue to predict mobility as a function of geographic distance.
Although alternatives exist to quantify \textit{functional} distances based on factors other than geography~\autocite{boschma2005proximity, brown1970functional, brown1970migration, kim2018functional}, geographic distance remains the unambiguous and foundational quantity for modeling mobility.

Here, we propose a neural embedding model that uses real-world mobility trajectories to encode the complex landscape of global mobility into a dense, continuous, and meaningful vector-space representation, from which a \textit{functional distance} between places can be derived.
This approach, origionally designed for learning language models~\autocite{mikolov2013word2vec} has been used to reveal insights in text~\autocite{linzhuo2020hyperbolic, tshitoyan2019mat2vec, garg2018gender, kozlowski2018geometry, hamilton2016diachronic, e2014doc2vec, nakandala2016twitch} and networks~\autocite{perozzi2014deepwalk, grover2016node2vec}.
In addition to providing a data-driven functional distance, this approach is computationally efficient, scalable for large datasets, robust to noise, and is readily interpretable through the application of simple vector arithmetic.

@@

The most fine-grained geographical unit of scientific production is the organization, such as an university, research institute, business, or government agency.
As a scientist's careers progress, they move between organizations or pick up additional simultaneous affiliations forming co-affiliation or \textit{mobility trajectories}.
These mobility trajectories, represented as ordered lists of organizations, can be derived from affiliations listed on researcher's publication histories~\autocite{robinson2019mobility, sugimoto2017mostimpact} (Fig.~\ref{fig:methods:mobility_sentence}).
We use the trajectories of the three million name-disambiguated authors who were \textit{mobile}---having more than one affiliation---between 2008 and 2019, as evidenced by their publications indexed in the Web of Science database.
A vector-space embedding of organizations is learned by using mobility trajectories as input to the standard skip-gram with negative sampling, or \textit{word2vec} neural-network architecture.
In the resulting embedding, each organization corresponds to a vector, and vectors relate to one another based on the likelihood of organizations appearing near together in the mobility trajectory data. 
The functional distance between two organizations is calculated as the cosine distance between their vectors, termed the \textit{embedding distance}.

@@

%
% Subsection: Applications
%
\subsection*{Embeddings capture global structure of mobility}

>>>>

Human mobility and migration drives major social phenomenon such as the growth and evolution of cities~\autocite{bettencourt2007cities, bettencourt2013cities, wef2017migration, curiel2018citygravity},  epidemics~\autocite{kraemer2020covid, truscott2012epidemicgravity, xia2005measlesgravity, hufnagel2004epidemics, eubank2004outbreaks}, economies~\autocite{kaluza2010cargo, kerr2011immigration, taylor2016refugees, pappalardo2015mobility}, innovation~\autocite{braunerhjelm2020labor, kaiser2018innovation, armano2017innovation, liu2010spillovers},  and scientific research~\autocite{sugimoto2017mostimpact, petersen2018multiscale, jonkers2013return, franzoni2014advantage, morgan2018prestige, rodrigues2016mobility}. 
Common models of mobility, such as the Gravity Model~\autocite{zipf1946gravity, xia2005measlesgravity, jung2008highwaygravity, barbosa2018human, curiel2018citygravity, lewer2008immigrationgravity, hong2016busgravity, truscott2012epidemicgravity, xia2005measlesgravity, simini2012universal}, describes mobility using the \textit{distance} between places.
However, amid a globalized world~\autocite{czaika2018migration, czaika2018globalisation, buchan2009globalization}, geographic distance has less relevance for understanding mobility, whereas other abstract notions of distance~\autocite{boschma2005proximity} pose practical and theoretical challenges.
Using the case of scientific mobility, we leverage the neural embedding model~\autocite{mikolov2013word2vec} to demonstrate a data-driven approach for learning a vector-space embedding that encodes a functional distance~\autocite{brown1970migration, brown1970functional, kim2018functional} between organizations and captures the latent and multi-faceted structure of human mobility.
We train neural embeddings of scientific organizations using mobility trajectories derived from the papers published between 2008 and 2019 of disambiguated authors from the Web of Science database.
Functional distance is derived as the cosine distance between vectors in the resulting embedding and is shown to better explain and predict real mobility between organizations than geographic distance. 
Exploiting their unique properties, we interpret the embedding space and observe how it captures latent factors related to culture, language, and prestige of organizations at multiple levels of granularity.
This approach can learn scalable, dense, and meaningful vector representations of scientific mobility directly from the data, and can easily be generalized to other mobility data.
\end{abstract}

@@

How far apart are two places?
The question is surprisingly hard to answer when it involves human mobility.
While geographic distance may have driven human movements in past eras, it holds less importance to a world connected by rapid transit, complex international relations, and global trade.
For instance, a person currently living in Australia is more likely to migrate to the United Kingdom, a country with similar language and culture, than to closer and more populous countries like Indonesia. 
Similarly, a prospective student living in South Korea is more likely to attend university in Canada rather than the closer though diplomatically isolated North Korea. 
Despite its complexity, ubiquitous models of human mobility such as the Gravity~\autocite{zipf1946gravity} and Radiation~\autocite{simini2012universal} model continue to predict mobility as a function of geographic distance.
Although alternatives exist to quantify \textit{functional} distances based on factors other than geography~\autocite{boschma2005proximity, brown1970functional, brown1970migration, kim2018functional}, geographic distance remains the unambiguous and foundational quantity for modeling mobility.

Here, we propose a neural embedding model that uses real-world mobility trajectories to encode the complex landscape of global mobility into a dense, continuous, and meaningful vector-space representation, from which a \textit{functional distance} between places can be derived.
This approach, originally designed for learning language models~\autocite{mikolov2013word2vec} has been used to reveal insights in text~\autocite{linzhuo2020hyperbolic, tshitoyan2019mat2vec, garg2018gender, kozlowski2018geometry, hamilton2016diachronic, le2014doc2vec, nakandala2016twitch} and networks~\autocite{perozzi2014deepwalk, grover2016node2vec}.
In addition to providing a data-driven functional distance, this approach is computationally efficient, scalable for large datasets, robust to noise, and is readily interpretable through the application of simple vector arithmetic.

@@

The most fine-grained geographical unit of scientific production is the organization, such as an university, research institute, business, or government agency.
As a scientist's careers progress, they move between organizations or pick up additional simultaneous affiliations forming co-affiliation or \textit{mobility trajectories}.
These mobility trajectories, represented as ordered lists of organizations, can be derived from affiliations listed on the researcher's publication histories~\autocite{robinson2019mobility, sugimoto2017mostimpact} (Fig.~\ref{fig:methods:mobility_sentence}).
We use the trajectories of the three million name-disambiguated authors who were \textit{mobile}---having more than one affiliation---between 2008 and 2019, as evidenced by their publications indexed in the Web of Science database.
A vector-space embedding of organizations is learned by using mobility trajectories as input to the standard skip-gram with negative sampling, or \textit{word2vec} neural-network architecture.
In the resulting embedding, each organization corresponds to a vector, and vectors relate to one another based on the likelihood of organizations appearing near together in the mobility trajectory data. 
The functional distance between two organizations is calculated as the cosine distance between their vectors, termed the \textit{embedding distance}.

@@

\subsection*{Embeddings capture global structure of mobility}
